Bootstrapping Multi-atlas Hippocampal Segmentation with \mb

### 2013-04-09 paper.Rnw
Results section DONE! Methods section DONE!  First draft anyway

bland_altman_plot <- function(x, y) {
  data = data.frame(x=x,y=y,diff=x-y)
  return(ggplot(data, aes(x= x, y = diff)) + 
    geom_hline(yintercept = mean(data$diff) + c(-2, 0, 2) * sd(data$diff), 
               linetype=2, color='brown') + 
    geom_point())
}
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Complete 1Yr Dataset Demographics Table                                  %
%                                                                             %
<<ADNI1-scr-demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
yr1 = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, 
               subset =  RID %in% yr1$RID & VISCODE %in% c('bl','m06','m12'),   
               method = "reverse", test=FALSE, overall=TRUE)

@@

\subsubsection{Schizophrenic First Episode Patients}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 15–30 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see Malla et
al. \citep{Malla2003}.

@@

%%%%%
%%%%%          Experiments: SZ First Episode Patient Validation           %%%%%
%%%%%
\subsubsection{SZ First Episode Patient Validation}

{\bf Dataset evaluated.} To validate that \mb performance generalises to
across disease categories, we measure the performance of \mb on the
Schizophrenic First Episode Patient (SFEP) dataset.

{\bf Atlas and template libary.} - two different atlas sets: a manual
hippocampal segmentation of patients, and Winterburn atlas set.  
\todo{complete}

{\bf Registration method.} \ants.
\todo{complete}

{\bf Label fusion.} Majority vote.
\todo{complete}

{\bf Evaluation.} We validate the FEP-atlas segmentations using Dice's Kappa,
and the Winterburn-atlas segmentations by correlating volumes. 
\todo{complete}

%%%%%
%%%%%              Experiments: Winterburn Atlases Validation              %%%%%
%%%%%
\subsubsection{Winterburn Atlases Validation}
{\bf Dataset evaluated.} - T1 BRAVO scans of the same subjects included in the 
Winterburn atlas set.  These scans are taken within .... weeks of the scans for 
the Winterburn atlases.
\todo{complete}

{\bf Atlas and template libary.} - Atlas library is Winterburn T1 atlases.  
Template library consists of all five T1 BRAVOs, plus 15 T1 healthy control images.
\todo{complete}

{\bf Registration method.} \ants
\todo{complete}

{\bf Label fusion.} Majority vote.
\todo{complete}

{\bf Evaluation.}  - Leave one out cross-validation (LOOCV) in which all five
subjects are segmented in separate runs of \mb.  In each run, the subject
to be segmented is excluded from the Atlas library (so only four atlases are used).

Segmentation accuracy is judged by difference in hippocampal volume.
\todo{complete}

@@

It is interesting to note that with an even number of templates, \mb shows a
small decrease in performance relative to when one fewer template image is used.
See section (\todo{ref}) for a discussion of this behaviour. In the remainder of
this section, only results from odd-sized template libraries will be shown.

@@

With an increasing number of templates, \mb-based segmentating using \ants
registration and majority vote label fusion shows improvement in overlap
accuracy over multi-atlas-based segmentation, using the same number of atlases
and voting method (Figure \todo{fig:ADNI1-xva-k-diff}. The magnitude of improvement over
multi-atlas-based segmentation decreases with an increasing number of atlases,
with accuracy converging with 7 atlases.  Peak improvement in \mb accuracy (~0.02
DSC) is found when one atlas is used with a template library of 20 images.
\todo{stats?}

@@

ggplot(stats.tests, 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases), 
      size = abs(statistic))) + 
  facet_grid( . ~ stat) + 
  geom_line(aes(size=0.1,area=0.1)) + 
  geom_point(aes(shape = factor(direction))) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_shape_discrete("t-statistic sign", labels=c("positive","negative")) + 
  scale_size_area(name="t-statistic magnitude") + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "p-value" ) +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb vs. multi-atlas segmentation 
  accuracy.}
  In these plots we consider the segmentation accuracy within each subject across
  the 10 rounds of validation per parameter setting. Standard deviation and 
  variance is computed per subject for \mb and multi-atlas, and the distribution
  of these test statistics is compared with a t-test. The p-value of this test, 
  is shown on the y-axis (scaled logarithmically).}
  \label{ADNI1-xval-variability}
\end{figure}

@@

Finally, \mb segmentation volumes show differential agreement with manual (SNT) 
volumes throughout the range of hippocampal volumes (Figure 
\ref{ADNI1-xval-Bland-Altman}). \mb shows a fixed bias towards producing smaller
segmentations, and shows a negligble proportional bias towards smaller 
segmentations inversely proportional to volume size. Nevertheless, when using 
3 or more atlases, \mb volumes are within the 95\% limits of agreement with 
manual labels.

\begin{figure}
<<ADN1-xval-Bland-Atlman, cache=TRUE, dependson='ADNI1-Xval-prep', fig.width=8, fig.height=4>>=
a_20_tmpls = subset(all_data, 
                      reg_method   == "ANTS" &
                      method.mb    == "Majority Vote" &
                      templates.mb == 20 &
                      atlases      %in% seq(1,9,by =2) &
                      volume.ma     < 10000)
total_HC_volume = ddply(a_20_tmpls, c("subject", "label", "atlases"), function (df) {
  data.frame(
    ma = sum(df$volume.ma),
    mb = sum(df$volume.mb)
  )
})
total_HC_volume$diff = total_HC_volume$ma - total_HC_volume$mb

ggplot(total_HC_volume, aes(x=ma, y = diff, group=atlases)) + 
    geom_hline(yintercept = mean(total_HC_volume$diff) + c(-2, 0, 2) * sd(total_HC_volume$diff), 
               linetype=2, color='brown') + 
    geom_point() + 
  facet_grid( . ~ atlases) + 
  xlab('Mean multi-atlas and MAGeT Brain total Hippocampus volume') + 
  ylab('Multi-atlas - MAGeT Brain volume') 
@
  \caption{Bland-Altman plot comparing multi-atlas and \mb hippocampal volumes 
  when using 9 atlases, and a 20 image template library.}
  \label{ADNI1-xval-Bland-Altman}
\end{figure}
% 
% 
% \begin{figure}
% <<ADN1-Xval-k-by-disease, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
% a_20_tmpls = subset(all_data, 
%                       reg_method  == "ANTS" &
%                       method.y    == "Majority Vote" &
%                       templates.y == 20 &
%                       atlases %in% c(1,9))
% template_library = read.csv('data/a2a_ants/template_library_composition.csv')
% template_library$subject = NULL
% a_20_tmpls = merge(a_20_tmpls, template_library,
%                    by.x=c('timestamp', 'templates.y'), 
%                    by.y=c('timestamp', 'templates'))
% ggplot(a_20_tmpls, aes(y=k.y, x=templates_CN,colour=DX)) +  # why does k.x look the same?
%   facet_grid( . ~ atlases) + 
%   geom_smooth(method='lm',fullrange=F) 
%  # add 
% @
%   \caption{Mean kappa vs. proportion of disease category in template library by disease category}
%   \label{}
% \end{figure}
% 
% 
% \todo{show cost (in registrations) / benefit tradec off graph:  show number of
% registrations per Kappa?  or hours of manual labour per Kappa?)}
% discuss run-time for each of MAGet and multi-atlas

@@

<<ADNI-src-volumes-prep,echo=F,cache=T>>=
bl    = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
mb    = read.csv("data/ADNI_baseline_volumes/ADNI1_1.5T_MAGeT_volumes.csv")
maper = read.csv("data/ADNI_baseline_volumes/MAPER_volumes.csv")
snt   = read.csv("data/ADNI_baseline_volumes/UCSFSNTVOL.csv")
fs    = read.csv("data/ADNI_baseline_volumes/UCSFFSX_02_15_12.csv")
fsl   = read.csv("data/ADNI_baseline_volumes/ADNI1_1.5T_FSLFIRST_volumes.csv", sep=';')
qc    = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_QC_Park.csv", sep="\t")

subjects = bl
subjects$VISCODE = 'bl' # to match up with adnimerge

@@

# diagnoses
dx = unique(adnimerge[c("DX.bl", "RID", "VISCODE")])

# Now create one data.frame with all the yr1 volume data we have from each datasource
combined = merge(subjects, mb_pruned   , by.x="Image.Data.ID", by.y="Source", all.x=TRUE)
combined = merge(combined, maper_pruned, by.x="Image.Data.ID", by.y="Source", all.x=TRUE)  # only Baseline
combined = merge(combined, snt_pruned  , by.x="Image.Data.ID", by.y="Source", all.x=TRUE)
combined = merge(combined, fs_pruned   , by.x="Image.Data.ID", by.y="Source", all.x=TRUE)
combined = merge(combined, fsl_pruned  , by.x="Image.Data.ID", by.y="Source", all.x=TRUE)
combined = merge(combined, dx          , by=c("RID","VISCODE"), all.x=TRUE)
combined = merge(combined, qc          , by=c("Image.Data.ID"), all.x=TRUE)

@@

Based on the results from the ADNI1 Cross-Validation experiment, in this
experiment \mb was configured with a template library of 21 randomly
chosen subject images (7 from each disease class) and used majority vote label
fusion. The entire ADNI1 Complete 1Yr 1.5T dataset was segmented by \mb and we 
now compare the resulting volumes with those obtained by manual segmentation (SNT), 
and other automated segmentation techniques (MAPER, FreeSurfer, and FSL).  Table
\ref{tab:ADN1-1.5T-Complete-1yr-Dataset-Segmentations} shows the total count of 
segmentations available, including a count of those which have failed a quality 
control inspection. Only those images which had segmentations from every method
are included in the following analysis (a total of  
\Sexpr{length(totals.complete$RID)} images).

@@

Comparing total hippocampal volume across segmentation methods, we find close
agreement between manual volumes and MAPER and MAGeT volumes across disease 
categories (Figure \ref{ADNI-scr-volumes-boxplot}). FSL and FreeSurfer both 
produce volumes which are demonstratively larger than manually segmented volumes.

@@

qplot(DX,value,data=melt(totals.complete,measure.vars=c("FS", "FSL", "MAPER","MAGeT","SNT")),
      colour=variable,geom="boxplot") + 
      xlab("Diagnosis") + ylab("Total Hippocampal Volume (mm3)")
@
  \caption{Comparison of hippocampus volumes obtained by FreeSurfer (FS), FSL, 
  MAPER, \mb (MAGeT) and manual (SNT).}
  \label{ADNI-scr-volumes-boxplot}
\end{figure}

% paired t-test against SNT to show similarity (should be low) 
% Kruskal-Willis against  (should be high?)
% post-hoc power analysis to say we can detect this change
\begin{figure}[h]
<<ADNI-scr-histogram,cache=T>>=
melted   = melt(totals.complete, measure.vars=c( "SNT","FSL","FS","MAPER","MAGeT"))
binwidth = 250

@@

melted.by.snt = melt(totals.complete, measure.vars=c("MAGeT","FSL", "FS", "MAPER"))
snt.tests <- ddply(melted.by.snt, c("variable", "DX"), function(df) {
  ks = ks.test(df$SNT, df$value)
  data.frame( 
    ks.D = ks$statistic,
    ks.p = ks$p.value
  )
})

label_x_pos = 10000
ggplot(data=melted, aes(x=value)) + 
  geom_histogram(binwidth=binwidth, alpha=0.8) + 
  geom_line(aes(y=freq, colour='red'), data=normaldf) +
  geom_text(aes(y=29,x=label_x_pos,hjust=1,size=1, 
                label=paste("list(W ==",signif(shapiro.w, digits=3), ", p == ", 
                            signif(shapiro.p,digits=3), ")")), 
            parse=TRUE, data=testsdf) +
  geom_text(aes(y=25,x=label_x_pos,hjust=1,size=1, 
                label=paste("list(D ==",signif(ks.D, digits=3), ", p == ", 
                            signif(ks.p,digits=3), ")")), 
            parse=TRUE, data=snt.tests) + 
  geom_text(aes(y=21,x=label_x_pos,hjust=1,size=1, 
                label=paste("sigma ==",signif(sd, digits=3))), 
            parse=TRUE, data=testsdf) + 
  geom_text(aes(y=17,x=label_x_pos,hjust=1,size=1, 
                label=paste("mu ==",signif(mean, digits=3))), 
            parse=TRUE,data=testsdf) + 
  xlim(0, 10000) + ylim(0,30) + 
  facet_grid(variable ~ DX) + 
  scale_colour_discrete(guide="none") + 
  scale_size_continuous(guide="none") + 
  labs(title="Total hippocampus volume by disease and measurement method", 
       x=expression(paste("Total hippocampal volume (", mm^2, ")")),
       y="Frequency") + 
  theme(axis.title.x = element_text(vjust=-0.1))
@
  \caption{{\bf ADNI Baseline cohort.} Comparison of total hippocampal volumes
  as measured by SNT, FSL, FreeSurfer (FS), MAPER, and \mb (MAGeT). A
  fitted normal curve is shown in red. $W$ is the Shapiro-Wilkes test statistic
  measuring normality of the data (signficance indicates non-normality). $D$ is
  the Kolmogorov-Smirnov test statistic measuring the goodness-of-fit between
  the distribution of measured volumes and SNT volumes (significance indicates a
  difference).}
  \label{ADNI-scr-histogram}
\end{figure}

%

@@

%%%%%
%%%%%             Results: First Episode Patient Validation          %%%%%
%%%%%
\subsection{First Episode Schizophrenic Patients}
High volume correlation between Winterburn segmentation volumes and ground
truth.  (High-ish?) Kappa when using manual segmentations as Atlases.

>>>>

bland_altman_plot <- function(x, y) {
  data = data.frame(x=x,y=y,diff=x-y)
  return(ggplot(data, aes(x= x, y = diff)) + 
    geom_hline(yintercept = mean(data$diff) + c(-2, 0, 2) * sd(data$diff), 
               linetype=2, color='brown') + 
    geom_point())
}

lm_eqn = function(m) {
  l <- list(a = format(coef(m)[1], digits = 2),
      b = format(abs(coef(m)[2]), digits = 2),
      r2 = format(summary(m)$r.squared, digits = 3));
  if (coef(m)[2] >= 0)  {
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2,l)
  } else {
    eq <- substitute(italic(y) == a - b %.% italic(x)*","~~italic(r)^2~"="~r2,l)    
  }

  as.character(as.expression(eq))
}
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1 Complete 1Yr Dataset Demographics Table                                  %
%                                                                             %
<<ADNI1-scr-demographics, echo=F, results="asis",cache=FALSE>>=
# RID < 2000 ensures only ADNI1 patients
yr1 = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
yr1$VISCODE <- factor(yr1$Visit, levels = c(1,2,3,4), labels=c("bl","m03","m06","m12"))
yr1 = subset(yr1, !grepl("Scaled_2", Description))
yr1 = yr1[,c("RID","VISCODE")]

adnimerge_yr1 = merge(adnimerge, yr1) 
tab <- summary(DX.bl ~ AGE + PTGENDER + PTEDUCAT + PTETHCAT + CDRSB + ADAS13 +
               MMSE, data = adnimerge, 
               subset =  adnimerge_yr1, #RID %in% yr1$RID & VISCODE %in% c('bl','m06','m12'),   
               method = "reverse", test=FALSE, overall=TRUE)

@@

\subsubsection{Schizophrenia First Episode Patients}
\label{sec:SZFEP-dataset}
All patients were recruited and treated through the Prevention and Early
Intervention Program for Psychoses (PEPP-Montreal), a specialized early
intervention service at the Douglas Mental Health University Institute in
Montreal, Canada. People aged 15–30 years from the local catchment area
suffering from either affective or non-affective psychosis who had not taken
antipsychotic medication for more than one month with an IQ above 70 were
consecutively admitted as either in- or out-patients. Of those treated at PEPP,
only patients aged 18 to 30 years with no previous history of neurological
disease or head trauma causing loss of consciousness were eligible for the
neuroimaging study; only those suffering from schizophrenia spectrum disorders
were considered for this analysis.  For complete program details see Malla et
al. \citep{Malla2003}.

@@

%%%%%
%%%%%          Experiments: SZ First Episode Patient Validation           %%%%%
%%%%%
\subsubsection{SZ First Episode Patient Validation}
The previous experiments explore \mb on an Alzheimer's disease dataset (ADNI).  To validate
that \mb algorithm works effectively with other diseased brain images, 
the performance of \mb is measured on a Schizophrenia dataset.

{\bf Dataset evaluated.} The entire Schizophrenia First Episode Patient (SZFEP) 
dataset (see section \ref{sec:SZFEP-dataset}).

{\bf Atlas and template libary.} The atlas library is composed of the Winterburn 
T1 atlas images, with whole hippocampus segmentations (no subregions). The 
template library is composed of 21 images selected at random from the SZFEP dataset.

{\bf Registration method.} The \ants registration algorithm is used for image
registration.

{\bf Label fusion.} The majority-vote method is used for label fusion.

{\bf Evaluation.} The manual segmentation protocol used to segment the Winterburn
atlases is similar to, but different from, the protocol used to segment the SZFEP
dataset. Therefore, rather than use an overlap metric, \mb hippocampal volumes
are compared to the corresponding manual segmentation volumes.

%%%%%
%%%%%              Experiments: Winterburn Atlases Validation              %%%%%
%%%%%
\subsubsection{Winterburn Atlases Cross-Validation}
In this experiment, the accuracy of the \mb algorithm on hippocampal subregion
segmentation is tested using a leave-one-out cross-validation (LOOCV) design.

{\bf Dataset evaluated.} \mb segmentations are evaluated on two datasets. The 
WA-BRAVO dataset consists of separately acquired 3T T1 BRAVO images (0.9mm-isotropic voxel)
of four of the five Winterburn atlas subjects. The WA-Resampled dataset consists
of the Winterburn atlas images and segmentations downsampled to 0.9mm-isotropic 
voxel resolution.

{\bf Atlas and template libary.} The atlas library is composed of the Winterburn 
T1 atlas images and hippocampal subregion segmentations. The template library 
consists of 15 3T T1 images (0.9mm-isotropic voxels) of healthy subjects as well
as all of the images from the dataset being evaluated.

\todo{reference, demographics for health control images}.

{\bf Registration method.} The \ants image registration method is used.

{\bf Label fusion.} The majority vote label fusion method is used.

{\bf Evaluation.}  Leave-one-out-cross-validation (LOOCV) is performed separately 
for each dataset evaluated (WA-BRAVO, and WA-Resampled) as follows. Each subject
in the dataset is segmented by \mb using an atlas library in which that subject's 
image is excluded. The segmentation volumes of each hippocampal subregion are
compared to the expert manual segmentations of the unmodified Winterburn atlases. 
As a point of comparison, we also calculate the subregion volumes from the Winterburn 
atlas segmentations after downsampling to 0.9mm-isotropic voxels.

@@

It is interesting to note that with an even number of templates, \mb shows a
small decrease in performance relative to when one fewer template image is used.
See section \todo{ref} for a discussion of this behaviour. In the remainder of
this section, only results from odd-sized template libraries will be shown.

@@

With an increasing number of templates, \mb-based segmentating using \ants
registration and majority vote label fusion shows improvement in overlap
accuracy over multi-atlas-based segmentation, using the same number of atlases
and voting method (Figure \ref{fig:ADNI1-xva-k-diff}). The magnitude of improvement over
multi-atlas-based segmentation decreases with an increasing number of atlases,
with accuracy converging with 7 atlases.  Peak improvement in \mb accuracy (~0.02
DSC) is found when one atlas is used with a template library of 20 images.

@@

ggplot(stats.tests, 
  aes(x=templates.mb, y=p.value, colour=as.factor(atlases), 
      size = abs(statistic))) + 
  facet_grid( . ~ stat) + 
  geom_line(aes(size=0.1,area=0.1)) + 
  geom_point(aes(shape = factor(direction))) + 
  geom_hline(aes(yintercept=0.05, alpha=0.5), linetype='dashed') + 
  geom_hline(aes(yintercept=0.01, alpha=0.5), linetype='dashed') + 
  scale_shape_discrete("t-statistic sign", labels=c("positive","negative")) + 
  scale_size_area(name="t-statistic magnitude") + 
  scale_x_continuous(breaks=seq(1,20,by=2)) + 
  scale_y_log10(limits=c(0.001,1), breaks=c(0.01,0.05,0.1,1.0)) + 
  xlab( "Number of Templates" ) + 
  ylab( "p-value" ) +
  scale_colour_hue(name="Atlases") +
  scale_alpha_continuous(guide='none') + 
  scale_linetype_discrete(guide='none') + 
  scale_size_continuous(guide='none') + 
  opts(legend.direction = "horizontal", legend.position = "bottom")
@
  \caption{{\bf Difference in variability of \mb vs. multi-atlas segmentation 
  accuracy.}
  Variability of segmentation accuracy within 10 rounds of validation per subject. 
  Standard deviation and variance is computed per subject for both \mb and 
  multi-atlas, and the distribution of these test statistics is compared with a 
  t-test. The p-value of this test, is shown on the y-axis (scaled logarithmically).}
  \label{ADNI1-xval-variability}
\end{figure}

@@

Finally, \mb segmentation volumes show differential agreement with manual (SNT) 
volumes throughout the range of hippocampal volumes (Figure 
\ref{ADNI1-xval-Bland-Altman}). \mb shows a negligble fixed bias towards 
producing smaller segmentations, and shows a slight proportional bias towards 
smaller segmentations with larger hippocampi.
\todo{Note limits of agreement?}

\begin{figure}
<<ADN1-xval-Bland-Atlman, cache=TRUE, dependson='ADNI1-Xval-prep', fig.width=8, fig.height=8>>=
df_subset = subset(all_data, 
                      reg_method   == "ANTS" &
                      method.mb    == "Majority Vote" &
                      templates.mb %in% c(1,11,19) &
                      atlases      %in% c(1,5,9) &
                      volume.ma     < 10000)

# get SNT data formatted like 
snt   = read.csv("data/a2a_snt_volumes.csv")
df_merged = merge(df_subset, snt, all.x = TRUE)

hc_vol = ddply(df_merged, c("subject", "label", "atlases", "templates.mb"), function (df) {
  data.frame(
    snt = sum(df$volume.snt),
    ma = sum(df$volume.ma),
    mb = sum(df$volume.mb)
  )
})
hc_vol$diff = hc_vol$snt - hc_vol$mb

group.stats = ddply(hc_vol, c("atlases", "templates.mb"), function (df) {
  data.frame(
    mean = mean(df$diff),
    sd   = sd(df$diff)
  )
})
hc_vol = merge(hc_vol, group.stats)
hc_vol$Atlases = hc_vol$atlases
hc_vol$Templates = hc_vol$templates.mb

ggplot(hc_vol, aes(x=ma, y = diff, group=Atlases)) + 
    geom_hline(aes(yintercept = mean + c(-2, 0, 2) * sd, colour='brown'), 
                linetype=2) + 
    geom_point() + 
    geom_smooth(method="lm") + 
  facet_grid(Atlases~Templates, labeller = label_both) + 
  xlab(expression('Mean manual and MAGeT-Brain total hippocampus volume ' (mm^3))) + 
  ylab(expression('Manual - MAGeT-Brain total hippocampus volume ' (mm^3))) 
@
  \caption{Bland-Altman plots comparing manual and \mb hippocampal volumes 
  when using a varying number of atlases and templates.}
  \label{ADNI1-xval-Bland-Altman}
\end{figure}
% 
% 
% \begin{figure}
% <<ADN1-Xval-k-by-disease, cache=TRUE, dependson='ADNI1-Xval-prep'>>=
% a_20_tmpls = subset(all_data, 
%                       reg_method  == "ANTS" &
%                       method.y    == "Majority Vote" &
%                       templates.y == 20 &
%                       atlases %in% c(1,9))
% template_library = read.csv('data/a2a_ants/template_library_composition.csv')
% template_library$subject = NULL
% a_20_tmpls = merge(a_20_tmpls, template_library,
%                    by.x=c('timestamp', 'templates.y'), 
%                    by.y=c('timestamp', 'templates'))
% ggplot(a_20_tmpls, aes(y=k.y, x=templates_CN,colour=DX)) +  # why does k.x look the same?
%   facet_grid( . ~ atlases) + 
%   geom_smooth(method='lm',fullrange=F) 
%  # add 
% @
%   \caption{Mean kappa vs. proportion of disease category in template library by disease category}
%   \label{}
% \end{figure}
% 
% 
% \todo{show cost (in registrations) / benefit tradec off graph:  show number of
% registrations per Kappa?  or hours of manual labour per Kappa?)}
% discuss run-time for each of MAGet and multi-atlas

@@

<<ADNI-src-volumes-prep,echo=F,cache=T>>=
#bl    = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
mb    = read.csv("data/ADNI_baseline_volumes/ADNI1_1.5T_MAGeT_volumes.csv")
maper = read.csv("data/ADNI_baseline_volumes/MAPER_volumes.csv")
snt   = read.csv("data/ADNI_baseline_volumes/UCSFSNTVOL.csv")
fs    = read.csv("data/ADNI_baseline_volumes/UCSFFSX_02_15_12.csv")
fsl   = read.csv("data/ADNI_baseline_volumes/ADNI1_1.5T_FSLFIRST_volumes.csv", sep=';')
qc    = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_QC_Park.csv", sep="\t")

# Get subjects, and diagnoses
yr1 = read.csv("data/ADNI_baseline_volumes/ADNI1_Complete_1Yr_1.5T_11_15_2012.csv")
yr1 = subset(yr1, !grepl("Scaled_2", Description)) #TODO: remove Scaled_2? 
yr1$VISCODE <- factor(yr1$Visit, levels = c(1,2,3,4), labels=c("bl","m03","m06","m12"))
subjects = yr1[,c("Image.Data.ID", "RID","VISCODE")]  # we only need these fields
dx = unique(adnimerge[c("DX.bl", "RID", "VISCODE")])
subjects = merge(subjects, dx)

@@

# Now create one data.frame with all the yr1 volume data we have from each datasource
combined = merge(subjects, mb_pruned   , by.x="Image.Data.ID", by.y="Source", all.x=TRUE)
combined = merge(combined, maper_pruned, by.x="Image.Data.ID", by.y="Source", all.x=TRUE)  # only Baseline
combined = merge(combined, snt_pruned  , by.x="Image.Data.ID", by.y="Source", all.x=TRUE)
combined = merge(combined, fs_pruned   , by.x="Image.Data.ID", by.y="Source", all.x=TRUE)
combined = merge(combined, fsl_pruned  , by.x="Image.Data.ID", by.y="Source", all.x=TRUE)
combined = merge(combined, qc          , by=c("Image.Data.ID"), all.x=TRUE)

@@

Based on the results from the ADNI1 Cross-Validation experiment, in this
experiment \mb was configured with a template library of 21 randomly
chosen subject images (7 from each disease class) and used majority vote label
fusion. The entire ADNI1 Complete 1Yr 1.5T dataset was segmented by \mb and we 
now compare the resulting volumes with those obtained by manual segmentation (SNT), 
and other automated segmentation techniques (MAPER, FreeSurfer, and FSL).  Table
\ref{tab:ADN1-1.5T-Complete-1yr-Dataset-Segmentations} shows the total count of 
segmentations available, including a count of those which have failed a quality 
control inspection. Only those images which had segmentations from every method
are included in the following analysis (a total of
\Sexpr{length(totals.complete$RID)} images; Table 
\ref{tab:ADNI1-1.5T-Complete-1Yr-Dataset-Method-Totals}).

@@

@@

qplot(DX,value,data=melt(totals.complete,measure.vars=c("FS", "FSL", "MAPER","MAGeT","SNT")),
      colour=variable,geom="boxplot") + 
      xlab("Diagnosis") + ylab("Total Hippocampal Volume (mm3)")
@
  \caption{Comparison of hippocampus volumes obtained by FreeSurfer (FS), FSL, 
  MAPER, \mb (MAGeT) and manual (SNT).}
  \label{fig:ADNI-scr-volumes-boxplot}
\end{figure}

% paired t-test against SNT to show similarity (should be low) 
% Kruskal-Willis against  (should be high?)
% post-hoc power analysis to say we can detect this change
\begin{figure}[h]
<<ADNI-scr-histogram-prep,cache=T>>=
melted   = melt(totals.complete, measure.vars=c( "SNT","FSL","FS","MAPER","MAGeT"))
binwidth = 250

@@

melted.by.snt = melt(totals.complete, measure.vars=c("MAGeT","FSL", "FS", "MAPER"))
snt.tests <- ddply(melted.by.snt, c("variable", "DX"), function(df) {
  ks = ks.test(df$SNT, df$value)
  data.frame( 
    ks.D = ks$statistic,
    ks.p = ks$p.value
  )
}) 
@

<<ADNI1-scr-histogram,dependson='ADNI-scr-histogram-prep',cache=T>>=

ggplot(data=melted, aes(x=value)) + 
  geom_histogram(binwidth=binwidth, alpha=0.8) + 
  geom_line(aes(y=freq, colour='red'), data=normaldf) +
  geom_text(aes(y=Inf,x=Inf,hjust=1.1,vjust=1.5,size=1, 
                label=paste("list(W ==",signif(shapiro.w, digits=3), ", p == ", 
                            round(shapiro.p,digits=3), ")")), 
            parse=TRUE, data=testsdf) +
  geom_text(aes(y=Inf,x=Inf,hjust=1.1,vjust=3,size=1, 
                label=paste("list(D ==",signif(ks.D, digits=3), ", p == ", 
                            round(ks.p,digits=3), ")")), 
            parse=TRUE, data=snt.tests) + 
  geom_text(aes(y=Inf,x=Inf,hjust=1.1,vjust=5.5,size=1, 
                label=paste("sigma ==",signif(sd, digits=3))), 
            parse=TRUE, data=testsdf) + 
  geom_text(aes(y=Inf,x=Inf,hjust=1.1,vjust=6,size=1, 
                label=paste("mu ==",signif(mean, digits=3))), 
            parse=TRUE,data=testsdf) + 
  #xlim(0, 10000) + 
  ylim(0,40) + 
  facet_grid(variable ~ DX) + 
  scale_colour_discrete(guide="none") + 
  scale_size_continuous(guide="none") + 
  labs(x=expression(paste("Total hippocampal volume (", mm^3, ")")),
       y="Frequency") + 
  theme(axis.title.x = element_text(vjust=-0.1))
@
  \caption{{\bf ADNI Baseline cohort.} Comparison of total hippocampal volumes
  as measured by SNT, FSL, FreeSurfer (FS), MAPER, and \mb (MAGeT). A
  fitted normal curve is shown in red. $W$ is the Shapiro-Wilkes test statistic
  measuring normality of the data (signficance indicates non-normality). $D$ is
  the Kolmogorov-Smirnov test statistic measuring the goodness-of-fit between
  the distribution of measured volumes and SNT volumes (significance indicates a
  difference).}
  \label{fig:ADNI-scr-histogram}
\end{figure}

Comparing total hippocampal volume, we find close
agreement between manual volumes, MAPER and MAGeT volumes across disease 
categories (Figure \ref{fig:ADNI-scr-volumes-boxplot}, Figure 
\ref{fig:ADNI-scr-histogram}). FSL and FreeSurfer both produce volumes in close 
agreement with each other, but which are consistently larger than manually 
segmented volumes in each disease category.

@@

%%%%%
%%%%%             Results: First Episode Patient Validation          %%%%%
%%%%%
\subsection{First Episode Schizophrenic Patients}
In this experiment we move from an Alzheimer's disease dataset to a dataset of 
first episode schizophrenia parients.  This dataset is segmented using \mb with the
Winterburn atlases, and a template library of 21 subject images selected at random. 
Expert manual whole hippocampal segmentations are used as gold standards.

\mb produces hippocampus segmentation volumes that are highly correlated with 
manual segmentation volumes (Figure \ref{fig:SZ-volumes}).  Additionally, as we 
saw in the previous experiments, \mb volumes show a fixed bias towards smaller 
volumes (although the bias is negligble), and a proportional bias towards 
producing smaller segmentations for larger hippocampi (Figure \ref{fig:FEP-SZ-Bland-Altman}.

\begin{figure}
<<FEP-volumes,cache=FALSE,dependson="setup">>=
SZ_volumes=read.csv("data/SZ_volumes_by_method.csv", sep="\t")
casted = dcast(SZ_volumes, Subject + Label ~ Method)
lm = lm(MAGeT ~ Manual, casted)
ggplot(data=casted, aes(x = Manual, y = MAGeT)) + 
  geom_smooth(method="lm", formula=y~x) + 
  geom_point() + 
  geom_text(aes(x=3500, y=5000, label=lm_eqn(lm),hjust=0,size=1), parse=TRUE, data=data.frame()) +
  xlab(expression("Manual left / right hippocampus volume " (mm^3))) + 
  ylab(expression("MAGeT left / right hippocampus volume "  (mm^3))) + 
  scale_size_continuous(guide="none") 
@
  \caption{{\bf First Episode Schizophrenic Patients.} Comparison of total
  HC volumes for \mb against manually rated Hippocampal volumes}
  \label{fig:SZ-volumes}
\end{figure}
